Genetic and karyotype divergence between parents affect clonality and sterility in hybrids

Version 1

Abstract

Abstract

Asexual reproduction can be triggered by interspecific hybridization, but its emergence is supposedly rare, relying on exceptional combinations of suitable genomes. To examine how genomic and karyotype divergence between parental lineages affect the incidence of asexual gametogenesis, we experimentally hybridized fishes (Cobitidae) across a broad phylogenetic spectrum, assessed by whole exome data. Gametogenic pathways generally followed a continuum from sexual reproduction in hybrids between closely related evolutionary lineages to sterile or inviable crosses between distant lineages. However, most crosses resulted in a combination of sterile males and asexually-reproducing females. Their gametes usually experienced problems in chromosome pairing, but females also produced a certain proportion of oocytes with premeiotically duplicated genomes, enabling their development into clonal eggs. Interspecific hybridization may thus commonly affect cell cycles in a specific way, allowing the formation of unreduced oocytes. The emergence of asexual gametogenesis appears tightly linked to hybrid sterility and constitutes an inherent part of the extended speciation continuum.

Introduction

Speciation is a continuous process, often progressing with decreasing gene flow between diverging populations. According to the concept of the speciation continuum 3, 4, hybridization between relatively closely related lineages typically results in fertile and recombining progeny and may even lead to an evolutionary meltdown or homoploid hybrid speciation 5. Increasing genetic divergence between hybridizing taxa may further promote the emergence of transgressive phenotypes 6 or hybrid vigour but typically lowers the mean fitness of hybrids, which consequently impacts the establishment of postzygotic reproductive isolating barriers 5. In species with well-differentiated sex chromosomes, incompatibilities particularly affect the heterogametic sex (Haldane’s rule) 7–9. Ultimately, speciation is completed when the accumulation of incompatible alleles in differentiating populations induces hybrid sterility or even inviability.

However, hybridization is also known to give rise to fertile, yet asexually reproducing hybrids in various animal and plant taxa. Asexual hybrids exhibit a broad spectrum of cytological mechanisms for the production of unreduced, often clonal, gametes, which range from entirely ameiotic processes (apomixis) to those involving more or less distorted meiosis (automixis) 10, 11. A relatively common gametogenic alteration leading to asexuality is premeiotic genome endoreplication (Fig. 1), found in a variety of hybrid fishes, amphibians, and reptiles 12–16. By auto-duplication, from each chromosome usually an identical copy is produced to pair with during the meiotic prophase. Premeiotic genome endoreplication thus not only ensures clonal reproduction, but also allows hybrids to overcome accumulated incompatibilities and problems in chromosome pairing that would otherwise lead to their sterility 15, 17. Hybridization and asexuality thus represent important evolutionary phenomena 10.

(a) Phylogenetic tree

based on exome-wide SNP data of the species used in the current crossing experiments. Red color indicates Bicanestrini group; green color indicates Cobitis sensu stricto group; yellow color indicates Adriatic group. (b) Distribution map of the spined loaches (Cobitis) included in this study. While Cobitis elongatoides, C. taenia, C. tanaitica, are known to act as parental species in hybridization events and emergence of clonal lineages, the other species (C. strumicae, C. ohridana, C. taurica, C. pontica) are known to be involved in secondary hybridization events. Abbreviations for all species in brackets. (c) Reproduction scheme of the clonal hybrids in the C. taenia complex, where E represents a haploid genome of C. elongatoides (orange), and T – of C. taenia (blue). Conventionally, hybridization between EE female and TT male lead to ET sterile males and clonal ET females that reproduce via gynogenesis. Gynogenetic females are pseudogamous, i.e. females produce diploid eggs via premeiotic genome endoreplication. Such eggs require sperm from a parental species (e.g. T sperm) to activate the embryonic development. The sperm genome is not incorporated in the hybrid’s genome but is instead eliminated after the egg’s activation, skipping karyogamy. All hybrid males are sterile because they do not produce spermatozoa or their extremely rare spermatozoa are aberrant and unable to fertilize eggs.

To explain the apparent link between asexuality and hybridization, the “balance hypothesis” has been proposed 18, which assumes that incompatibilities, accumulated among parental genomes, may disturb gametogenesis in hybrids leading to the formation of unreduced gametes. Hybrid asexuality may thus emerge in a “favourable evolutionary window” of genetic divergence between hybridizing species, which should be large enough to trigger particular gametogenic aberrations, but no too large to compromise the hybrid’s fertility or even viability. The recently proposed concept of the “extended speciation continuum” 9 explicitly links the emergence of asexual reproduction to the speciation continuum. While forming species ultimately reach complete intrinsic reproductive isolation under increased divergence, fertile hybrids may be produced at earlier stages that exhibit aberrant gametogenesis, leading towards clonal reproduction. This not only facilitates the formation of allodiploid and allo-polyploid asexual hybrid lineages, but asexuality per se also effectively restricts introgression between parental species due to non-recombinant production of gametes. Hybrid asexuality may thus be considered a particular type of Bateson-Dobzhansky-Muller (BDM)-incompatibilities, representing a form of intrinsic reproductive isolation during the speciation process 17.

While the evolutionary link between hybridization and asexuality has been empirically documented, it remains unclear how widespread such a link is and how commonly might asexual gametogenesis occur in hybrids. On the one hand, naturally occurring asexual hybrids exist in many major animal and plant lineages, but appear rare, suggesting that the genomic pre-conditions required for asexual reproduction may be rarely met during hybridization 19. On the other hand, most of the knowledge about the evolution and biology of asexual hybrids comes from studies of naturally occurring lineages, while laboratory crossing experiments have been rarely performed to directly address the rate of clonal gametogenesis in hybrids. The questions how easily asexual reproduction is induced and why it is so often linked with a hybrid constitution thus remains unclear.

To estimate the incidence of clonal gametogenesis in a radation of freshwater fish, we performed extensive crossing experiments of sexual species and investigated the gametogenic pathways in their hybrids, using European loaches of the family Cobitidae as a model. Cobitidae presents a speciose group of freshwater fish of South Asian origin that spread over most of the Palearctic region since late Eocene 20. They colonized Europe in three distinct lineages 21, which diverged from each other around 17-20 Mya 21, 22. The so-called Adriatic and Bicanestrinia lineages colonized the Balkan and Anatolian regions, while Central and Eastern Europe have been colonized by the Cobitis sensu stricto lineage (Fig. 1a-b). This lineage is composed of two sub-groups, the so-called C. taenia clade, involving several closely related species diverged during the last 1 Mya, and C. elongatoides, which diverged ∼9 Mya from the C. taenia clade 17. Such independent colonization events and subsequent range shifts gave rise to natural hybrid zones across the Palearctic and Europe in particular 23. Diploid and polyploid hybrids have been reported within the family, including gynogenetic ‘asexual’ forms, which produce clonal eggs but require the fertilization with a sperm from sexual males in order to trigger their development (Fig. 1a-c). Such forms usually discard the sperms ’genome after fertilization and are currently found in hybrids of the Asian Misgurnus anguillicaudatus complex 24, the Cobitis hankugensis-longicorpa hybrids in Korea 25, and in hybrids of the European Cobitis sensu stricto lineage (Fig. 1b), 26. In these natural systems, only hybrid females are known to produce unreduced gametes, employing premeiotic genome endoreplication, while hybrid males are usually sterile due to aberrant pairing of chromosomes in meiosis 15, 27.

In the present paper, we report meiotic and premeiotic gametogenic stages of experimental hybrids of eight loach species, representing the three main European lineages. These species have been crossed in various combinations from closely related taxa to phylogenetically distant ones, which allowed us to test, how often unreduced hybrid gametes arise, whether their emergence is linked to genetic and karyotypic divergence between parental species, and how asexuality is related to hybrid sterility.

Results

Parental species and their karyotypes

The selected parental species represent three distinct phylogenetic lineages from Europe (Fig. 1a, Supplementary Fig. S1) 21, possessing diverse karyotypes with diploid chromosome sets (2n) between 48 – 50 chromosomes and variable numbers of meta-/submeta- and subtelo-/acrocentrics (Supplementary Fig. S2, Supplementary Table S1). Karyotype dissimilarity among our focal species, measured as the autosomal karyotype index (AKD) 28, significantly increases with their genetic divergence, measured by the occurrence of heterozygous SNPs (Mantel test, r = 0.338, p = 0.0261). However, the correlation was not always linear (Supplementary Fig. S3). This is because some phylogenetically distant species belonging to the Bicanestrinia and Adriatic phylogroups possess morphologically almost identical karyotypes (Supplementary Fig. S2, Supplementary Table S1).

Artificial crosses and the presence of germ cells in all F1 hybrids

We crossed the eight focal species in 12 combinations, including some reciprocal crosses (Fig. 2). To investigate the effect of increasing phylogenetic distance, we predominantly used EE males and females for crosses with most species belonging to the three sampled phylogroups, but species combinations without C. elongatoides were performed too. We obtained viable progeny of both sexes from all attempted species combinations, altogether resulting in 19 F1 hybrid families (Fig. 2, Supplementary Table S2 Supplementary Fig. S1). We observed no significant deviations from equal sex ratios among the F1 (Binomial test, all p values > 0.25 after corrections for multiple testing; Supplementary Table S2) and - about 6 months after hatching - randomly selected 3-5 juveniles per family to examine the gametogenesis in F1 hybrids.

Matrix with successfully established and analyzed crosses between eight species of Cobitis.

Each cell represents a particular cross between parental species including the information about maternal and paternal species (i.e. the direction of the cross). Phylogenomic tree plotted along margins indicates the relationships among crossed species based on exome-wide SNP data. Each color represents a particular reproductive output in F1 hybrids; green labeling indicates production of haploid gametes via normal meiosis; yellow color indicates the F1 progeny which produce unreduced gametes and present fully developed gonads; Red color denotes sterile progeny, predominantly referring to hybrid males. Grey labels potentially sterile hybrids for which we did not obtain diplotenic cells to fully confirm their reproductive output. The striped cell represents the F1 hybrid combination which was previously described 17 and involves the species used in this study.

Whole-mount immunofluorescence staining of gonads with antibodies against Vasa protein showed that the analysed F1 hybrids contained germ cells organized in clusters, which were similar to germ cell clusters of sexual fish, used as a control group (Supplementary Fig.s S4a,c,e). The immunofluorescence staining of SYCP3 protein was performed either on whole-mount staining of gonadal tissue or on pachytenic spreads, and revealed the formation of lateral components of synaptonemal complexes (SCs) in most controls and F1 hybrids, suggesting the presence of meiocytes (Fig. 3,4 Supplementary Fig. S4b,d,f, S7, S9, S10). The only exceptions were PS and TB hybrid juvenile males, whose gonads contained clusters of germ cells without meiocytes. The absence of meiocytes was later verified also in both, adult PS and TB males (Supplementary Fig. S4a,b).

Comparison of pachytene spermatocytes between male of sexual diploid species (a), hybrids’ genomotypes (b,c), and corresponding gametogenic pathways (d-f).

The spread of pachytene spermatocytes of C. elongatoides with 25 completely paired bivalents (a), and diploid NP hybrid with almost paired chromosomes (b) and OE hybrid with aberrant pairing, including bivalents and univalents (c). Thick arrows indicate bivalents; thin arrows indicate univalents. Scale bar = 10 µm. Schematic representation of the gametogenic pathway including presumptive karyotype composition in gonocytes and pachytene cells in males of sexual species (d) and NP hybrids (e) that are able to complete pairing and form gametes meiotically, and OE hybrids (f) which exhibit abnormal pairing leading to sterility.

Comparison of pachytene oocytes between female of sexual diploid species (a) and hybrids genomotypes (b,c,d) and corresponding gametogenic pathways (e-h).

The SC spreads from pachytene oocyte of C. ohridana with fully paired 25 bivalents (a), diploid OB hybrid with almost paired chromosomes (b) and OE hybrid female which have pachytene cells with aberrant pairing (c) and cells with fully paired chromosomes emerged after premeiotic genome endoreplication (d). Thick arrows indicate bivalents; thin arrows indicate univalents. Scale bar = 10 µm. Schematic representation of gametogenic pathway including presumptive karyotype composition in gonocytes and pachytene cells. Females of sexual species (e) and OB hybrid (f) which are able to fully or partially complete pairing of chromosomes and form gametes meiotically; OE hybrids (g) exhibit two populations of pachytene oocytes: oocytes with unduplicated genomes and oocytes with duplicated genome. Oocytes with unduplicated genome (g, left) have abnormal pairing leading to the inability of proceed beyond pachytene and thus sterility. Oocytes with duplicated genomes (g, right) have normal pairing and thus leading to the formation of unreduced gametes.

Analysis of meiocytes

We then investigated chromosome pairing during the pachytene stage of the meiotic prophase by immunostaining of the SYCP3 and SYCP1 components of the synaptonemal complex, and we analysed lampbrush chromosomes of the diplotene meiotic stage in females.

Sexual species produce reduced gametes with properly formed bivalents

Depending on their chromosome numbers, parental species always contained 24-25 synaptonemal complexes (Fig. 3a; Fig. 4a). Flow cytometry of testicular cell suspensions from parental species revealed the presence of haploid and diploid cells with a major peak corresponding to haploid cells (Supplementary Fig. S5a). Histological analysis showed many sperm cells organized in cysts in all sexual males investigated, containing properly developed Sertoli and Leydig cells (Supplementary Fig. S6a), suggesting normal development of testes.

Sterility of hybrid males is caused by both, the aberrant pairing in the meiotic prophase and the inability to proceed to meiosis

Flow cytometry revealed that the analysed hybrid males lacked the haploid population of cells, characteristic for spermatids and spermatozoa of sexual males. Hybrid males had only diploid cell populations (Supplementary Fig. S5b-h). Interestingly, TB males had a small population of diploid cells corresponding to pachytene stage. These males had morphologically underdeveloped gonads, indicating that here the sterility is probably caused by the inability of gonocytes to proceed to meiosis (Supplementary Fig. S4a,b; Supplementary Fig. S5h). Histological examinations demonstrated that analysed hybrid males, had defective testes with an asynchronous development of germinal cells in cysts (Supplementary Fig. S6b-g). Similar results were reported in naturally occurring ET hybrid males 15.

Furthermore, hybrid males from most experimental crosses showed aberrant chromosomal pairing during the pachytene, with variable numbers of univalents, bivalents and multivalents, (Fig. 3c; Supplementary Fig. S7a-g). The only exception to aforementioned observations were the NP hybrids involving closely related parental species (C. tanaitica and C. pontica) where we detected fully-paired chromosomes during the pachytene (Fig. 3b). Chromosomal spreads of NP males also clearly showed the presence of formed spermatid nuclei, similar to sexual males (Fig. 3b). Additionally, we counted SCs, based on the presence of crossing-over foci (Supplementary Fig. S10a-f) 15, 29.

Production of unreduced gametes by premeiotic genome endoreplication is the prevalent gametogenic pathway among hybrid females

Analysis of diplotenic oocytes of most hybrid females revealed no univalents or mispaired chromosomes, but instead showed the formation of 49 or 50 properly formed bivalents (Supplementary Fig. S8a-e). Similar to previous data on natural clones 13, 27, we therefore conclude that diplotene oocytes in the majority of hybrid females contain only cells with duplicated genomes, suggesting their ability to complete clonal gametogenesis (Fig. 4; Supplementary Fig. S8a-e). The only exceptions were observed in OB and NP hybrids, where diplotenic oocytes contained only 25 fully formed bivalents, suggesting normal meiosis with the pairing likely between orthologous chromosomes (Supplementary Fig. S8f).

However, the analysis of pachytenic oocytes revealed different patterns. While OB and NP females possessed 25 bivalents with delayed pairing of two chromosome pairs (Supplementary Fig. S9d), generally matching their diplotene configurations, most F1-females contained pachytenic cells with abnormally-paired chromosomes and various numbers of univalents, bivalents, and multivalents. (Fig. 4c; Supplementary Fig. S9a-c, e-h). This suggests that most pachytenic oocytes in hybrids did not undergo genome endoreplication. Pachytenic oocytes with duplicated genomes and thus normal chromosomal pairing were only observed in females from three F1-families (EC, EO and ET hybrids). These showed sets of 50 or 49 bivalents with properly loaded lateral and central synaptonemal components and MLH1 loci on each bivalent (Fig. 4d).

Overall, the incidence of genome endoreplication was very low among pachytenic cells of any F1 hybrid female, occurring on average in 0.7% of cells only (Supplementary Table S4) and there were no significant differences between any combinations of parental species (Wald Z-test of generalized linear mixed-effect mode (GLMM); DF = 9, p-value = 0.903). We also did not observe any significant effects of maternal species in those crosses, where both directions were performed, i.e. there were no obvious differences between ET and TE as well as between OE and EO cross types.

We therefore treated all experimental progenies as a single group and tested whether they differed from previously-analysed natural diploid ET-hybrids, whose incidence of duplicated gonocytes was higher, on average ∼6%; 27. Here, the difference was significant (Wald Z-test of GLMM; DF = 1; p-value = 0.0067), suggesting that natural clones produce a higher proportion of oocytes with endoreplicated genomes than experimetal F1 hybrids.

The contrasting incidence of oocytes with endoreplicated genomes between pachytene (aprox 6%) and diplotene stages (100%) is consistent with previous results 27 suggesting that premeiotic genome endoreplication is a rare event in all types of crosses. This allows cells to form normal bivalents and to enter the diplotene, while the majority of cells remain non-duplicated and cannot proceed beyond the pachytene checkpoint due to mispairing among orthologous chromosomes.

Divergence of parental karyotypes affects the formation of bivalents between orthologous chromosomes in hybrids but has different effects in males and females

We than focused exclusively on those pachytenic cells with unduplicated genomes and found that F1-hybrids largely vary in numbers of paired orthologous chromosomes, ranging from 3 detectable synaptonemal complexes in EO males to 25 fully-formed bivalents in OB females. The observed numbers of bivalents in pachytenic cells were negatively correlated with morphological dissimilarity of parental karyotypes expressed by the AKD index (GLMM z value = -6.71; p < 10-10). Males tended to have fewer bivalents than females (z = -3.07; p = 0.002) and we noticed that interaction with hybrids’ sex was significant too (z = -2.68; p = 0.007) with a steeper slope in males (-0.097) (Fig. 5). This suggests that chromosomal dissimilarity between parental species negatively impacts the formation of bivalents in their hybrids, and that male offspring generally form less bivalents and tend to be more adversely affected by divergence of parents than female progeny (significant interaction with sex).

Effect of karyotype dissimilarity on numbers of bivalents in hybrids.

Boxplots showing the number of synaptonemal complexes per cell (y-axis) of the F1 hybrids ranked along x-axis according to the morphological divergence of karyotypes between parental species expressed by AKD index 28. Numbers indicate the AKD index values, subscripts F and M indicate females and males, respectively and experimental crosses are indicated by respective letter combinations.

Comparison with other crossing experiments within Cobitidae

Finally, to put our results in context with previously published data, we plotted the reproductive outputs of all known loach hybrids in relation to genetic divergence and chromosomal differences between their parental species. To express the amount of chromosomal dissimilarity among parental species, we used the AKD index calculated from published and here newly prepared karyotypes of all spined loaches known to hybridize. To express the amount genetic divergence among these species, we used their sequence distances in the nuclear RAG1 gene, as a proxy, because exome-based sequence data are not available for some of the previously examined crosses. Data on RAG1- and karyotype divergences were then correlated with reproductive outputs in their hybrids and presented on Fig. 6a.

Reproductive outcomes of hybrids from crosses between species from the subfamily Cobitinae (a).

Plot demonstrating the relationship between reproductive outcome and viability of F1 hybrid loaches and the genetic and karyotype divergences among parental species. Data on F1 hybrids have been obtained in this study as well as from the literature. Karyotype divergence is marked as AKD index and genetic distance is estimated from published RAG1 sequences (K2P distance). Green color shows the ranges of hybrids with sexual reproduction; yellow color shows the ranges of F1 hybrids, which produce unreduced gametes; purple color indicates the ranges of sterility in both sexes; red color shows the ranges of inviable hybrids. To rank reproductive outcomes for every type of cross we labeled them with additional symbols: – both sexes producing haploid gametes with fully paired chromosomes; –both sexes are fertile producing haploid gametes, some females produce unreduced gametes; – females produce haploid eggs, males are sterile; – females produce clonal eggs, males are sterile; – both sexes are sterile; – both sexes are inviable. Asterisk (*) indicates cross combinations for which diplotene analysis is missing. Each abbreviation represent a subgenome of the parental species: Cobitis elongatoides = EE; C. taenia = TT; C. tanaitica = NN; C. taurica = CC; C. pontica = PP; C. strumicae = SS; C. bilineata = BB; C. ohridana = OO; C. hankugensis = HH; C. biwae = BW; C. lutheri = Lut; C. tetralineata = Tet; Iksookimia longicorpa = LL; M. anguillicaudatus clade A = MaA; M. anguillicaudatus clade B = MaB; Lefua echigonia = Lef. Hypothetical scheme of the selective processes necessary for the emergence of clonality in F1 hybrids (b). First step includes the accumulation of mutations in gametogenic regulatory genes which is required to induce premeiotic genome endoreplication in F1 hybrid females. Second step includes the chromosomal divergence which leads to the aberrant chromosomal pairing followed by the cell arrest of meiocytes with non-duplicated genomes. Premeiotic genome endoreplication is sex specific, while hybrid males are sterile.

Inviable or fully sterile hybrids were generally found among hybrids between the genetically most distant species, while sexual and/or clonal reproduction were observed in hybrids between more closely related species. However, the window of genetic divergences allowing for sexual as well as asexual reproduction was relatively wide (Fig. 6a). Interestingly, we noticed that asexuality tends to be observed among female hybrids between species with substantially diverged karyotypes, while sexual reproduction generally required more similar karyotypes.

Discussion

Using freshwater fishes, loaches (Cobitidae), as a model, the present study reveals a wide range of reproductive outcomes in F1 hybrids in relation to genetic and karyotype divergences between the parental species (Fig. 6). On one extreme, hybrid inviability has been reported in crosses between distant species 30. On the other extreme, the production of haploid gametes was found in crosses of closely related species, such as between C. lutheri and C. tetralineata 31, and between C. pontica and C. taenia 17 or C. tanaitica species, whose divergence has been dated to ∼1 Mya 17, but also in OB female hybrids in a relatively distant species pair (Fig. 2, Fig. 6, Supplementary Fig. S9d, S10d,). However, the most common types of reproductive outputs were hybrid sterility and production of unreduced gametes.

In case of hybrids between relatively far-distant species pairs, like the intergeneric Cobitis-Misgurnus cross, or between Adriatic and Cobitis sensu stricto lineages (i.e. EB and TB hybrids), sterility occurred in hybrids of both sexes (Fig. 6). However, the most common outcome was asymmetric, with sterility affecting hybrid males, while F1 females produced unreduced oocytes. This pattern occurred in the majority of crosses, including species pairs, like SE, SP or EO, with divergences reaching ∼17 or even >20 Mya back 22 (Fig. 2, Fig. 6) and it has also been reported from several natural hybrid complexes 13, 15, 27. An intermediate situation was observed in TP-hybrids between a closely related species that diverged ∼1 Mya 17, where males produced reduced gametes and females produced both, reduced and unreduced ones.

It is generally assumed that the emergence of asexuality is a rare phenomenon that may require complex changes in gametogenic pathways and reproductive modes 19. This opinion might be corroborated by the “twiggy” distribution of ‘asexual’ organisms on the tree of life 32 and by many failures to re-generate naturally occurring asexual hybrids by experimental crosses of their parental species (e.g. in apomictic Poecilia fish and parthenogenetic Darevskia lizards 19, 33. However, we observed the production of unreduced gametes in 7 out of 11 crossed species pairs of loaches and in 10 out of 18 species pairs when including literature reports and natural hybrids (Fig. 2, Fig. 6). Such a widespread emergence of asexual gametogenesis therefore appears not to depend on the presence of a genome from any particular parental species, but to arise relatively commonly among independent species pairs across the entire subfamily Cobitinae. Asexual gametogenesis has also been observed in experimental F1 progeny of several other vertebrate species 34. This suggests that the formation of unreduced gametes, i.e. the necessary prerequisite for asexual reproduction, may be a much more common outcome of interspecific hybridization than it would appear from the rarity of asexual hybrid taxa that are successfully established in nature.

Some studies also suggested that the emergence of asexuality requires the evolution of specific genes within hybrid lineages after their origin 35. However, our finding that asexual gametogenesis originates relatively commonly already in the F1 generation somewhat refines this hypothesis and suggests that it is a direct result of genome merging by interspecific hybridization. Importantly, all F1-females capable of clonal gametogenesis employed an identical cytogenetic mechanism, i.e. premeiotic genome endoreplication, as also observed in many natural asexual vertebrates 13, 15, 16, 27, 34, 36, 37 suggesting that gametogenic perturbations induced by hybridization are often canalised into a similar developmental pathway. The reasons for such common patterns are not clear. The regulation of cell cycle is particularly sensitive to stress and perturbations 38 and interspecific hybridization may act as such stimulus 18. While asexual reproduction may employ various pathways 10, it is possible that aberrations of cell cycle leading to endoreplication are more common since they require less accumulated incompatibilities than other routes to the production of unreduced gametes. Alternatively, the occurrence of endoreplication may reflect some predispositions, like the spontaneous production of unreduced gametes in some sexual parental lineages, which would then ensure clonal reproduction in their hybrids. However, our data argue against this alternative, since we analysed hundreds of gonocytes of sexual females without a single incidence of genome endoreplication 15, 27. Moreover, during years of crossing experiments and population sampling, we genotyped hundreds of C. elongatoides and C. taenia and their progenies and never found an autotriploid specimen 23, 39, which would be expected if fertilisation of unreduced gametes existed in these species 40.

Our in-depth analysis of hybrid gametogenesis in Cobitis loaches further provided compelling evidence for a tight link between hybrid incompatibility, asexuality and sterility. Namely, the majority of meiocytes produced by both the male and female hybrids failed to form a full set of bivalents between orthologous chromosomes and were arrested at meiotic checkpoints (Fig. 3). However, unlike their sterile brothers, hybrid F1 females also usually contained a minor proportion of oogonia that underwent premeiotic genome endoreplication. In most our F1 hybrid females, only these rare oogonia were able to complete meiosis and form diploid gametes. Interestingly, similar observations were reported from several natural and experimental hybrid vertebrates, like ET and EN hybrid loaches, geckos and whiptail lizards (Aspidoscelis) Medaka ricefish (Oryzias) 16, 27, 34, 41. This suggests that even successfully established hybrid asexuals usually contain only a minor proportion of oogonia, which actually undergo clonal gametogenetic pathway.

Such congruent patterns across phylogenetically distant taxa thus conform to the concept of an “extended speciation continuum”. Similar to the classical concept of the “speciation continuum”, it assumes that speciation generally proceeds from early stages characterized by the production of sexual hybrid progeny to irreversibly diverged species, whose hybrids are either sterile or even inviable. At intermediate stages, however, hybrids between diverging evolutionary lineages may not only have a certain incidence of sterile gonocytes, but also tend to produce unreduced gametes (Fig. 6). Genome endoreplication thus appears as an effective way to overcome sterility in hybrid females, even at high levels of divergence. Asexuality may thus often represent the only reproductive pathway of hybrids between a given pair of species and, together with hybrid sterility, it may simultaneously contribute to reproductive isolation of parental species due to clonal genome propagation.

The fact that emergence of clonality is sex-specific and may rescue fertility in females, while males of the same cross usually remain sterile, indicates a potential role of incompatibilities between sex chromosomes (Haldane’s rule) or other uniparentally inherited factors (Darwin’s corollary 43. Unfortunately, the sex determination systems are not yet known in most of the investigated species 9 but the reasons for such an asymmetry are likely even more complex. For instance, meiocytes in hybrid males contained significantly lower numbers of properly-formed synaptonemal complexes than non-duplicated meiocytes in females (Fig. 5) and were more severely affected by decreasing similarity of parental karyotypes (see Results, 2.3.5.). This suggests that the pairing mechanisms and affinity among orthologous chromosomes fundamentally differ between males and females, e.g. 44. Moreover, we recently found that spermatogonial cells taken from a sterile ET hybrid male may resume the ability of endoreplication when transplanted into female gonads 39. This suggests that initiation of premeiotic endoreplication may rely on female-specific gonadal environment rather than on genetic sex determination.

The genetic divergence among parents may affect early gametogenic stages in hybrids through various mechanisms causing e.g. depletion of primordial germ cells 45, improper pairing of diverged chromosomes 46, asynapsis between sex chromosomes 47, or deleterious epistatic interactions between orthologous genes and their binding motifs on chromosomes 48. In loaches, we documented at least two such mechanisms underlying hybrid sterility. TB and PS hybrid males had almost no spermatocytes, suggesting their gonocytes probably failed the transition to meiosis. The other type of sterility conforms to a chromosomal speciation model 49, 50, prevailing in most F1 hybrid males, in which spermatocytes developed but possessed aberrantly-paired chromosomes, preventing further progression to spermatids. Analogous patterns were observed in non-duplicated oocytes of most hybrid females, suggesting that problems in bivalent-formation have a detrimental impact on the meiosis in both male and female loach hybrids.

Karyotype evolution among parental species indeed plays a crucial role in hybrids’ reproductive capabilities 50. Our data suggest that it has a crucial, yet indirect, impact on asexual reproduction. Specifically, hybrids between species with diverged karyotypes had less bivalents in non-duplicated meiocytes than those combining similar karyotypes. Sometimes, even hybrids between phylogenetically distant, yet karyotypically similar species could produce full sets of bivalents (e.g. OB hybrids) (Fig. 5). On the other hand, the numbers of endoreduplicated oocytes were low in any hybrid. Their proportion was relatively higher in natural clones than in experimental F1 Cobitis, suggesting that such a trait may recover during the evolutionary establishment of a natural asexual lineage 27 and/or that selection among “freshly” emerged clones would favour those with the highest proportion of endoreduplicated oocytes. Nevertheless, even the most widespread clones possessed only ∼6% of such oogonia, suggesting that karyotype similarity allows the completion of meiosis and hybrids would mostly reproduce sexually by reduced gametes, even if they had the capacity of producing unreduced gametes, too.

While hybridization is assumed to present an important trigger of asexuality in vertebrates and other organisms 10, the origins of clonal lineages are thought to be rare, and often can be traced back to a single or few hybridization events in a given species pair 19, 26. Multiple independent origins of clonal gametogenesis in spined loaches 23, 26, this study) in part contradicts this view since interspecific hybridization has been found to result in clonal gametogenesis relatively commonly up to a substantial divergence, where fertile or viable hybrids are no longer appearing (Fig. 6a). However, we emphasize that successful establishment of hybrid clones under natural conditions is likely restricted by several additional levels of selective pressure and thus filtering (Fig. 6b). On the one hand, the “balance hypothesis” 18 and its expansions and fine-tuned versions 17, 19, suggest that the production of unreduced gametes probably requires some level of genetic divergence among the parental species to alter the hybrid cell cycle towards genome endoreplication. On the other hand, it appears that most meiocytes of asexual hybrids anyway do not pass through endoreplication 16, 27, 34, 41. Consequently, if the parental karyotypes remain largely similar, then orthologous chromosomes may successfully pair even in such nonduplicated meiocytes, potentially, allowing them to accomplish normal meiosis and form reduced gametes. This would decrease the proportion of unreduced gametes produced by such hybrids and negatively impact on their likelihood to establish a stable asexual lineage.

We therefore propose, that for a cross to be successful in generating an asexual hybrid, it not only has to produce a sufficient number of unreduced gametes, but it also has to prevent the formation of reduced gametes from non-duplicated meiocytes. One way to achieve this may be if hybridization involves species with sufficiently diverged karyotypes, ensuring improper pairing of orthologous chromosomes in non-duplicated gametes. Increased divergence of parental karyotypes may therefore have a positive impact on the production of stable clonal hybrids and, for instance, the fixation of certain chromosomal rearrangements among diverging populations may be accelerated by certain geographical aspects of speciation, such as parapatry 49, 50. It would thus be attractive to test if the incidence of asexual hybrids correlates with increased rates of karyotype evolution among parental species.

The rarity of naturally occurring clonal organisms probably results from the simultaneous impact of the abovementioned selective processes. However, our data indicate that the production of unreduced gametes, which is the very basic conditio sine qua non for the evolution of asexuality, may be surprisingly “easy” to trigger, especially in hybrids of distantly related lineages. Of course, we may not rule out that the frequent emergence of premeiotic endoreplication in Cobitidae reflects some specific predispositions in their gametogenic machineries in comparison to other taxa, where asexuality appears rarer. Thus, our study demonstrates that conclusions about the evolution of asexuality should not only be drawn from natural asexuals, but should be based also on experimental crosses.

eLife Assessment

This important paper provides new insight into how asexual reproduction can arise in interspecific hybrids, an empirical example of the "extended speciation continuum." The molecular cytogenetic evidence supporting the conclusions is solid, but additional replicates within each species would strengthen the overall findings. By highlighting the potential for asexuality to evolve in hybrids during a narrow window of species divergence, this work will be of broad interest to evolutionary biologists.

Reviewer #1 (Public Review):

This paper provides new evidence on the relationship between genetic/chromosome divergence and capacity for asexual reproduction (via unreduced, clonal gametes) in hybrid males or females. Whereas previous studies have focussed just on the hybrid combinations that have yielded asexual lineages in nature, the authors take an experimental approach, analysing meiotic processes in F1 hybrids for combinations of species spanning different levels of divergence, whether or not they form asexual lineages in nature. As such, the findings here are a substantial advance towards understanding how new asexual lineages form.

The quality of the work is high, the analyses are sound, and the authors sensibly link their observations to the speciation continuum. I should also add that the cytogenetic work here is just beautiful!

A key finding is that the precondition for asexual reproduction - the formation of unreduced gametes - is not unusual among hybrid females, so that we have to consider other factors to explain the rarity of asexual species - a major unresolved issue in evolutionary biology. This work also highlights a previously overlooked effect of chromosome organisation on speciation.

Reviewer #2 (Public Review):

The authors investigate the origin of asexual reproduction through hybridization between species. In loaches, diploid, polyploid, and asexual forms have been described in natural populations. The authors experimentally cross multiple species of loaches and conduct an impressively detailed characterization of gametogenesis using molecular cytogenetics to show that although meiosis arrests early in male hybrids, a subset of cells in females undergo endoreplication before meiosis, producing diploid eggs. This only occurred in hybrids of parental species that were of intermediate divergence. This work supports an expanding view of speciation where asexuality could emerge during a narrow evolutionary window where genomic divergence between species is not too high to cause hybrid inviability, but high enough to disrupt normal meiotic processes.

I enjoyed reading this study and I appreciate the amount of work it takes to conduct these types of cytogenetic experiments. But, my main concern with this study is I was left wondering if the sample sizes are large enough to get a sense how variable endoreplication is in these loach species. Most of the hybrids between species are the result of crosses between 1-2 families. Within males and females, meiocyte observations are limited to a handful of pachytene and diplotene stages. I think it would be helpful to be more transparent about the sample sizes in the main text.

Along these lines, the authors argue against the possibility that endoreplication may be predisposed to occur at a higher rate in some species (line 291). Instead, they suggest that endoreplication is a result of perturbing the cell cycle by combining the genomes of two different species. Their main argument is based on gonocyte counts from parental females in a previous reference. It is essential to include counts from the parents used in this study to make a clear comparison with the F1s.

In the discussion (lines 320-333), the authors postulate the sex-specific clonality they observe could be a result of Haldane's rule. Given these fish do not have known sex chromosomes, I do not find this argument strong. Haldane's rule refers to the exposure of recessive incompatibilities with the sex chromosomes in the hybrid heterogametic sex. This effect would therefore be limited to degenerated sex chromosomes where much of the sequence content on the Y or W has been lost. These species may have homomorphic sex chromosomes, but if this is the case, they likely are not very degenerated. Instead, it seems more plausible that the sex-specific effect the authors observe is due to intrinsic differences of spermatogenesis and oogenesis. Is there any information about sex-specific differences in the fidelity of gametogenesis from other species that would support a higher likelihood of endoreplication?

The final thing I was left wondering about was this missing link between endoreplication and activating embryonic development of the diploid egg. In these loach species, a sperm is required to activate egg development, but the sperm genome is discarded (line 100). What is the mechanism of this and how does it evolve concurrently during hybridization?

Author Response

The following is the authors’ response to the original reviews.

Thank you very much for the kind comments about our manuscript. We have improved the text to address all reviewers’ comments and suggestions. Additionally, we corrected and improved the supplementary tables.

Reviewer #1 (Public Review):

This paper provides new evidence on the relationship between genetic/chromosome divergence and capacity for asexual reproduction (via unreduced, clonal gametes) in hybrid males or females. Whereas previous studies have focussed just on the hybrid combinations that have yielded asexual lineages in nature, the authors take an experimental approach, analysing meiotic processes in F1 hybrids for combinations of species spanning different levels of divergence, whether or not they form asexual lineages in nature. As such, the findings here are a substantial advance towards understanding how new asexual lineages form.

The quality of the work is high, the analyses are sound, and the authors sensibly link their observations to the speciation continuum. I should also add that the cytogenetic work here is just beautiful!

A key finding is that the precondition for asexual reproduction - the formation of unreduced gametes - is not unusual among hybrid females, so that we have to consider other factors to explain the rarity of asexual species - a major unresolved issue in evolutionary biology. This work also highlights a previously overlooked effect of chromosome organisation on speciation.

Thank you for the nice comments about our work as well as for appreciating our cytogenetics work and figures.

Reviewer #2 (Public Review):

The authors investigate the origin of asexual reproduction through hybridization between species. In loaches, diploid, polyploid, and asexual forms have been described in natural populations. The authors experimentally cross multiple species of loaches and conduct an impressively detailed characterization of gametogenesis using molecular cytogenetics to show that although meiosis arrests early in male hybrids, a subset of cells in females undergo endoreplication before meiosis, producing diploid eggs. This only occurred in hybrids of parental species that were of intermediate divergence. This work supports an expanding view of speciation where asexuality could emerge during a narrow evolutionary window where genomic divergence between species is not too high to cause hybrid inviability, but high enough to disrupt normal meiotic processes.

Thank you.

I enjoyed reading this study and I appreciate the amount of work it takes to conduct these types of cytogenetic experiments. But, my main concern with this study is I was left wondering if the sample sizes are large enough to get a sense how variable endoreplication is in these loach species. Most of the hybrids between species are the result of crosses between 1-2 families. Within males and females, meiocyte observations are limited to a handful of pachytene and diplotene stages. I think it would be helpful to be more transparent about the sample sizes in the main text.

Thank you for raising this point. We have improved the Supplementary Tables S2 and S3 to clarify how many individuals we analyzed from each genetic family and added this information to the main text. In total we obtained 12 combinations with 19 F1 hybrid families. For the combination, C. elongatoides x C. taenia hybrids we obtained three families, for C. elongatoides x C. ohridana, C. elongatoides x C. tanaitica, C. elongatoides x C. bilineata and C. ohridana x C. bilineata, we obtained two families For the rest of the combinations of hybrids we obtained single family. From these families, 79 individuals were used for the analysis of the meiocites. Additionally, 24 parental individuals, males and females, were analysed. For the parental species, we analysed 852 cells, for hybrid males we investigated 244 cells, and 665 cells for hybrid females.

Along these lines, the authors argue against the possibility that endoreplication may be predisposed to occur at a higher rate in some species (line 291). Instead, they suggest that endoreplication is a result of perturbing the cell cycle by combining the genomes of two different species. Their main argument is based on gonocyte counts from parental females in a previous reference. It is essential to include counts from the parents used in this study to make a clear comparison with the F1s.

Thank you, we agree with your comment and included the observations of meiocytes from several parental species, i.e. C. elongatoides, C. taenia, C. pontica, C. tanaitica, and C. ohridana. Among 852 cells analyzed, we did not observe cells with duplicated genomes and abnormalities in chromosomal pairing. By contrast, among 665 pachytene cells of F1 hybrid females, we revealed altogether ~1% of endoreplicated ones. We tested these data by binomial GLM and found these differences to be significant, suggesting that sexuals, even if they may have some unnoticed duplication events, clearly have a significantly lower incidence of abnormal pachytene cells. We have now included this information in the main text.

In the discussion (lines 320-333), the authors postulate the sex-specific clonality they observe could be a result of Haldane's rule. Given these fish do not have known sex chromosomes, I do not find this argument strong. Haldane's rule refers to the exposure of recessive incompatibilities with the sex chromosomes in the hybrid heterogametic sex. This effect would therefore be limited to degenerated sex chromosomes where much of the sequence content on the Y or W has been lost. These species may have homomorphic sex chromosomes, but if this is the case, they likely are not very degenerated. Instead, it seems more plausible that the sex-specific effect the authors observe is due to intrinsic differences of spermatogenesis and oogenesis. Is there any information about sex-specific differences in the fidelity of gametogenesis from other species that would support a higher likelihood of endoreplication?

Thank you for this important question, however, we think it was a misunderstanding. We do not postulate that our observation conforms to Haldanes’ rule as, by contrast to this rule based on sex chromosomes, our previous publication demonstrated that whatever the gonadal sex differentiation is in our taxa, the ability to overcome sterility by asexual gametogenesis is always confined to female gonadal environment (or oogenesis in general), even in the transplanted spermatogonial cells (Tichopad et al. 2022). What we meant by our text is that our results do not fully conform to Haldane’s rule. We therefore reworded our text to rule out such a misconception.

Nonetheless, we note that it has been demonstrated that Haldanes’ rule is also applicable to species with little differentiated sex chromosomes (e.g. Presgraves and Orr 1998) and that recessive incompatibilities are not the only explanation as faster male theory or faster X may also apply in such cases (Dufresnes et al. 2016). Therefore, we have kept our remarks about Haldane’s rule here. Moreover, for several parental species, we preliminary found the occurrence of an XY gonadal sex differentiation system, albeit these are unpublished and need further validation.

The final thing I was left wondering about was this missing link between endoreplication and activating the embryonic development of the diploid egg. In these loach species, a sperm is required to activate egg development, but the sperm genome is discarded (line 100). What is the mechanism of this and how does it evolve concurrently during hybridization?

Thank you for the comment. There have been many speculations about why gynogens actually need sperm to activate their egg development, but to our knowledge, no explanation has been validated to date. Interestingly, a recent theoretical model by Fyon et al. BiorXiv 2023 suggested that the ability of sperm exclusion may evolve separately from the ability to produce clonal eggs. Hence, this topic is complex and remains unresolved, and we feel that it is out of the scope of the present MS. We have slightly modified the text and added 2 refs., to address your suggestion.

Reviewer #1 (Recommendations For The Authors):

The paper is well prepared - though the resolution of Fig 1 on the pdf is rather poor.

Thank you! We have now provided the high-resolution figures.

Overall, I have few suggestions for improvements:

Line 58. How does endoduplication itself "overcome accumulated incompatibilities" other than failure of synapsis? Perhaps by maintaining the F1 state, and so avoiding reduced fitness arising from recombination and disruption of coadapted gene combinations.

We have added a sentence to the main text “Premeiotic genome endoreplication thus not only ensures clonal reproduction but also allows hybrids to overcome problems in chromosome pairing that would otherwise lead to their sterility 15,17.” that we hope sufficiently addresses this issue.

Line 118 - please explain the AKD index here - as you have some in SI. Also please be clearer on how you measure genetic divergence as proportion of heterozygous SNPs - presumably this is via exon sequences from F1 females?

Please note that we have explained the AKD index in the relevant part of the Methods section already. However, we have now also added a brief explanation to the Results section, as suggested. We apologize for imprecise description of the genetic divergence measurements. As described in the Methods section, this is not measured by heterozygosity (as we wrongly stated here), but as p-distance among sequences of coding regions between parental species.

Lines 126 ff. It is unfortunate that the design of the crosses was not more balanced or extensive. Nonetheless, I do appreciate the effort involved here and think the results are solid as is.

Thank you.

Line 142. Please define PS and TB (and other acronyms) at first use.

We have added the definition for all acronyms at the first use.

Lines 192-193. What about EP and EN - as shown to have unreduced gametes in Fig. 2?

Thank you for this question. Based on analyses of the diplotene stage, we showed that EP and EN hybrids produced diploid eggs. However, in pachytene, we did not find duplicated oocytes due to the rarity of endoreplication. Similarly, the low incidence of duplicated pachytene cells was observed in natural as well as F1-hybrids in loaches and reptiles (Newton et al., 2016, Dedukh et al., 2021, 2022).

Lines 217-219. The observed correlation of chromosome divergence (AKD index) and numbers of bivalents in pachytene makes sense and is an important observation. Did this GLM simultaneously consider the effect of genetic divergence (as implied in methods)?

Thank you for this comment. We originally tested separately the fit of two models, one with AKD and the other with SNP divergence. Since the AKD model significantly outperformed the SNP-based one, we focused our interpretation on the former. However, as you suggested, we now re-calculated the model taking into account the joint effects of both predictors in a single model and indeed, this model outperformed both single predictors. In conclusion, while AKD is still the strongest single predictor for the observed amounts of bivalents, the additional effect of genetic distance still significantly improves the model fit. We have now included this result into the main text.

This finding does not alter our conclusions, it just suggests that the effect of chromosomal morphology is probably more complex, involving the role of more subtle sequence divergence or structural variants.

Line 242. The Discussion is a great read - careful interpretation and a really interesting interpretation in context of the broader literature.

Thank you for the appreciation. Your positive feedback and evaluation are highly motivating us to expand our work.

Line 396. Some references from book chapters (18, 52) are incomplete. Please fix.

We have now corrected these references accordingly.

Reviewer #2 (Recommendations For The Authors):

Transparency about meiocyte sample sizes: These counts are all in supplemental table 3. From this table, it is unclear if a majority of these meiocytes are from a single individual or from multiple males or females. Or, in the crosses where there are multiple families, are the meiocytes sampled from all families? I am trying to get a sense whether endoreplication and the fidelity of oogenesis could be influenced by genetic variants segregating within species. If the meiotcytes are only sampled from a single individual from a single cross, you may not see this variation. If this is the case, perhaps the correlation between genetic divergence and the formation of asexual clones may not be as strong. Additional replicates may not be feasible, but at a minimum I think it would be helpful to address whether endoreplication could or could not be variable and if the sample sizes are sufficient.

Thank you for raising this point. We have improved the Supplementary table to clarify how many individuals we analyzed from each family and added this information to the main text. Unfortunately, additional replicates are not feasible due to the long generation time of the fish. We otherwise agree with your comment and included this point in the Discussion.

Gonocyte counts from parental females: The authors say they "analysed hundreds of gonocytes of sexual females without a single incidence of genome endoreplication." I could not find a clear count in the references given. They note that the incidence of endoreplication was very low in pachytene cells in this study (0.7%).

Thank you, we agree with your comment and included the observations of meiocytes from several parental species, i.e. C. elongatoides, C. taenia, C. pontica, C. tanaitica, and C. ohridana. Among 852 cells analyzed, we did not observe cells with duplicated genomes and abnormalities in chromosomal pairing. By contrast, among 665 pachytenic cells of F1 hybrid females, we revealed altogether ~1% of endoreplicated ones. We tested these data by binomial GLM and found these differences to be significant, suggesting that sexuals, even if they may have some unnoticed duplication events, clearly have significantly lower incidence. of abnormal pachytene cells. We have now included this information in the main text.

They refer to supplemental table 4 (line 196), which does not exist in the supplement. The authors should report these numbers in the revised manuscript.

Thank you for pointing this out. We have corrected the name of the supplementary table, it actually is supplementary table S3.

Version 2

Abstract

Abstract

Asexual reproduction can be triggered by interspecific hybridization, but its emergence is supposedly rare, relying on exceptional combinations of suitable genomes. To examine how genomic and karyotype divergence between parental lineages affect the incidence of asexual gametogenesis, we experimentally hybridized fishes (Cobitidae) across a broad phylogenetic spectrum, assessed by whole exome data. Gametogenic pathways generally followed a continuum from sexual reproduction in hybrids between closely related evolutionary lineages to sterile or inviable crosses between distant lineages. However, most crosses resulted in a combination of sterile males and asexually-reproducing females. Their gametes usually experienced problems in chromosome pairing, but females also produced a certain proportion of oocytes with premeiotically duplicated genomes, enabling their development into clonal eggs. Interspecific hybridization may thus commonly affect cell cycles in a specific way, allowing the formation of unreduced oocytes. The emergence of asexual gametogenesis appears tightly linked to hybrid sterility and constitutes an inherent part of the extended speciation continuum.

Introduction

Speciation is a continuous process, often progressing with decreasing gene flow between diverging populations. According to the concept of the speciation continuum 3,4, hybridization between relatively closely related lineages typically results in fertile and recombining progeny and may even lead to an evolutionary meltdown or homoploid hybrid speciation 5. Increasing genetic divergence between hybridizing taxa may further promote the emergence of transgressive phenotypes 6 or hybrid vigour but typically lowers the mean fitness of hybrids, which consequently impacts the establishment of postzygotic reproductive isolating barriers 5. In species with well-differentiated sex chromosomes, incompatibilities particularly affect the heterogametic sex (Haldane’s rule) 7–9. Ultimately, speciation is completed when the accumulation of incompatible alleles in differentiating populations induces hybrid sterility or even inviability.

However, hybridization is also known to give rise to fertile, yet asexually reproducing hybrids in various animal and plant taxa. Asexual hybrids exhibit a broad spectrum of cytological mechanisms for the production of unreduced, often clonal, gametes, which range from entirely ameiotic processes (apomixis) to those involving more or less distorted meiosis (automixis) 10,11. A relatively common gametogenic alteration leading to asexuality is premeiotic genome endoreplication (Fig. 1), found in a variety of hybrid fishes, amphibians, and reptiles 12–16. By auto-duplication, from each chromosome usually an identical copy is produced to pair with during the meiotic prophase. Premeiotic genome endoreplication thus not only ensures clonal reproduction, but also allows hybrids to overcome problems in chromosome pairing that would otherwise lead to their sterility 15,17. Hybridization and asexuality thus represent important evolutionary phenomena 10.

(a) Phylogenetic tree

based on exome-wide SNP data of the species used in the current crossing experiments. Red color indicates Bicanestrini group; green color indicates Cobitis sensu stricto group; yellow color indicates Adriatic group. (b) Distribution map of the spined loaches (Cobitis) included in this study. While Cobitis elongatoides, C. taenia, C. tanaitica, are known to act as parental species in hybridization events and emergence of clonal lineages, the other species (C. strumicae, C. ohridana, C. taurica, C. pontica) are known to be involved in secondary hybridization events. Abbreviations for all species in brackets. (c) Reproduction scheme of the clonal hybrids in the C. taenia complex, where E represents a haploid genome of C. elongatoides (orange), and T – of C. taenia (blue). Conventionally, hybridization between EE female and TT male lead to ET sterile males and clonal ET females that reproduce via gynogenesis. Gynogenetic females are pseudogamous, i.e. females produce diploid eggs via premeiotic genome endoreplication. Such eggs require sperm from a parental species (e.g. T sperm) to activate the embryonic development. The sperm genome is not incorporated in the hybrid’s genome but is instead eliminated after the egg’s activation, skipping karyogamy. All hybrid males are sterile because they do not produce spermatozoa or their extremely rare spermatozoa are aberrant and unable to fertilize eggs.

To explain the apparent link between asexuality and hybridization, the “balance hypothesis” has been proposed 18, which assumes that incompatibilities, accumulated among parental genomes, may disturb gametogenesis in hybrids leading to the formation of unreduced gametes. Hybrid asexuality may thus emerge in a “favourable evolutionary window” of genetic divergence between hybridizing species, which should be large enough to trigger particular gametogenic aberrations, but no too large to compromise the hybrid’s fertility or even viability. The recently proposed concept of the “extended speciation continuum” 9 explicitly links the emergence of asexual reproduction to the speciation continuum. While forming species ultimately reach complete intrinsic reproductive isolation under increased divergence, fertile hybrids may be produced at earlier stages that exhibit aberrant gametogenesis, leading towards clonal reproduction. This not only facilitates the formation of allodiploid and allopolyploid asexual hybrid lineages, but asexuality per se also effectively restricts introgression between parental species due to non-recombinant production of gametes. Hybrid asexuality may thus be considered a particular type of Bateson-Dobzhansky-Muller (BDM)-incompatibilities, representing a form of intrinsic reproductive isolation during the speciation process 17.

While the evolutionary link between hybridization and asexuality has been empirically documented, it remains unclear how widespread such a link is and how commonly might asexual gametogenesis occur in hybrids. On the one hand, naturally occurring asexual hybrids exist in many major animal and plant lineages, but appear rare, suggesting that the genomic pre-conditions required for asexual reproduction may be rarely met during hybridization 19. On the other hand, most of the knowledge about the evolution and biology of asexual hybrids comes from studies of naturally occurring lineages, while laboratory crossing experiments have been rarely performed to directly address the rate of clonal gametogenesis in hybrids. The questions how easily asexual reproduction is induced and why it is so often linked with a hybrid constitution thus remains unclear.

To estimate the incidence of clonal gametogenesis in a radiation of freshwater fish, we performed extensive crossing experiments of sexual species and investigated the gametogenic pathways in their hybrids, using European loaches of the family Cobitidae as a model. Cobitidae presents a speciose group of freshwater fish of South Asian origin that spread over most of the Palearctic region since late Eocene 20. They colonized Europe in three distinct lineages 21, which diverged from each other around 17-20 Mya 21,22. The so-called Adriatic and Bicanestrinia lineages colonized the Balkan and Anatolian regions, while Central and Eastern Europe have been colonized by the Cobitis sensu stricto lineage (Fig. 1a-b). This lineage is composed of two sub-groups, the so-called C. taenia clade, involving several closely related species diverged during the last 1 Mya, and C. elongatoides, which diverged ∼9 Mya from the C. taenia clade 17. Such independent colonization events and subsequent range shifts gave rise to natural hybrid zones across the Palearctic and Europe in particular 23. Diploid and polyploid hybrids have been reported within the family, including gynogenetic ‘asexual’ lineages. These lineages reproduce via gynogenesis, i.e. clonal eggs require sperm from sexual males to activate further development into genetically identical progeny. After fertilization, male pronucleus is usually eliminated from the eggs, without genetic contribution to the progeny 24,25 (Fig. 1a-c). Such forms usually discard the sperms ‘genome after fertilization and are currently found in hybrids of the Asian Misgurnus anguillicaudatus complex 26, the Cobitis hankugensis-longicorpa hybrids in Korea 27, and in hybrids of the European Cobitis sensu stricto lineage (Fig. 1b), 28. In these natural systems, only hybrid females are known to produce unreduced gametes, employing premeiotic genome endoreplication, while hybrid males are usually sterile due to aberrant pairing of chromosomes in meiosis 15,29.

In the present paper, we report meiotic and premeiotic gametogenic stages of experimental hybrids of eight loach species, representing the three main European lineages. These species have been crossed in various combinations from closely related taxa to phylogenetically distant ones, which allowed us to test, how often unreduced hybrid gametes arise, whether their emergence is linked to genetic and karyotypic divergence between parental species, and how asexuality is related to hybrid sterility.

Results

Parental species and their karyotypes

The selected parental species represent three distinct phylogenetic lineages from Europe (Fig. 1a, Supplementary Fig. S1) 21, possessing diverse karyotypes with diploid chromosome sets (2n) between 48 – 50 chromosomes and variable numbers of meta-/submeta- and subtelo-/acrocentrics (Supplementary Fig. S2, Supplementary Table S1). Karyotype dissimilarity among our focal species was measured as the autosomal karyotype index (AKD) which evaluates the differences in chromosome numbers and fundamental numbers of their arms 30. We found that AKD significantly increases with the genetic divergence among analyzed parental species, evaluated as the p-distances in SNPs in coding sequences (Mantel test, r = 0.338, p = 0.0261). However, the correlation was not always linear (Supplementary Fig. S3). This is because some phylogenetically distant species belonging to the Bicanestrinia and Adriatic phylogroups possess morphologically almost identical karyotypes (Supplementary Fig. S2, Supplementary Table S1).

Artificial crosses and the presence of germ cells in all F1 hybrids

We crossed the eight focal species in 12 combinations, including some reciprocal crosses (Fig. 2). To investigate the effect of increasing phylogenetic distance, we predominantly used EE males and females for crosses with most species belonging to the three sampled phylogroups, but species combinations without C. elongatoides were performed too. We obtained viable progeny of both sexes from all attempted species combinations, altogether resulting in 19 F1 hybrid families (Fig. 2, Supplementary Table S2 Supplementary Fig. S1). Here we present the abbreviation for every hybrid combination: C. elongatoides × C. taenia (ET), C. elongatoides × C. tanaitica (EN), C. elongatoides × C. taurica (EC), C. elongatoides × C. pontica (EP), C. elongatoides × C. strumicae (ES), C. elongatoides × C. bilineata (EB), C. elongatoides × C. ohridana (EO), C. taenia × C. bilineata (TB), C. ohridana × C. bilineata (OB), C. tanaitica × C. pontica (NP), C. pontica x C. strumicae (PS). We observed no significant deviations from equal sex ratios among the F1 (Binomial test, all p values > 0.25 after corrections for multiple testing; Supplementary Table S2) and -about 6 months after hatching -randomly selected 3-5 juveniles per family to examine the gametogenesis in F1 hybrids.

Matrix with successfully established and analyzed crosses between eight species of Cobitis.

Each cell represents a particular cross between parental species including the information about maternal and paternal species (i.e. the direction of the cross). Phylogenomic tree plotted along margins indicates the relationships among crossed species based on exome-wide SNP data. Each color represents a particular reproductive output in F1 hybrids; green labeling indicates production of haploid gametes via normal meiosis; yellow color indicates the F1 progeny which produce unreduced gametes and present fully developed gonads; Red color denotes sterile progeny, predominantly referring to hybrid males. Grey labels potentially sterile hybrids for which we did not obtain diplotenic cells to fully confirm their reproductive output. The striped cell represents the F1 hybrid combination which was previously described 17 and involves the species used in this study.

Whole-mount immunofluorescence staining of gonads with antibodies against Vasa protein showed that the analysed F1 hybrids contained germ cells organized in clusters, which were similar to germ cell clusters of sexual fish, used as a control group (Supplementary Fig. S4a,c,e). The immunofluorescence staining of SYCP3 protein was performed either on whole-mount staining of gonadal tissue or on pachytenic spreads, and revealed the formation of lateral components of synaptonemal complexes (SCs) in most controls and F1 hybrids, suggesting the presence of meiocytes (Fig. 3,4 Supplementary Fig. S4b,d,f, S7, S9, S10). The only exceptions were PS and TB hybrid juvenile males, whose gonads contained clusters of germ cells without meiocytes. The absence of meiocytes was later verified also in both, adult PS and TB males (Supplementary Fig. S4a,b).

Comparison of pachytene spermatocytes between male of sexual diploid species (a), hybrids’ genotypes (b,c), and corresponding gametogenic pathways (d-f).

The spread of pachytene spermatocytes of C. elongatoides with 25 completely paired bivalents (a), and diploid NP hybrid with almost paired chromosomes (b) and OE hybrid with aberrant pairing, including bivalents and univalents (c). Thick arrows indicate bivalents; thin arrows indicate univalents. Scale bar = 10 µm. Schematic representation of the gametogenic pathway including presumptive karyotype composition in gonocytes and pachytene cells in males of sexual species (d) and NP hybrids (e) that are able to complete pairing and form gametes meiotically, and OE hybrids (f) which exhibit abnormal pairing leading to sterility.

Comparison of pachytene oocytes between female of sexual diploid species (a) and hybrids genotypes (b,c,d) and corresponding gametogenic pathways (e-h).

The SC spreads from pachytene oocyte of C. ohridana with fully paired 25 bivalents (a), diploid OB hybrid with almost paired chromosomes (b) and OE hybrid female which have pachytene cells with aberrant pairing (c) and cells with fully paired chromosomes emerged after premeiotic genome endoreplication (d). Thick arrows indicate bivalents; thin arrows indicate univalents. Scale bar = 10 µm. Schematic representation of gametogenic pathway including presumptive karyotype composition in gonocytes and pachytene cells. Females of sexual species (e) and OB hybrid (f) which are able to fully or partially complete pairing of chromosomes and form gametes meiotically; OE hybrids (g) exhibit two populations of pachytene oocytes: oocytes with unduplicated genomes and oocytes with duplicated genome. Oocytes with unduplicated genome (g, left) have abnormal pairing leading to the inability of proceed beyond pachytene and thus sterility. Oocytes with duplicated genomes (g, right) have normal pairing and thus leading to the formation of unreduced gametes.

Analysis of meiocytes

We then investigated chromosome pairing during the pachytene stage of the meiotic prophase by immunostaining of the SYCP3 and SYCP1 components of the synaptonemal complex, and we analysed lampbrush chromosomes of the diplotene meiotic stage in females. We analysed 24 parental individuals, including males and females (EE, TT, NN, PP, OO) and 79 individuals among all F1 families (Supplementary Tables S2, S3).

Sexual species produce reduced gametes with properly formed bivalents

Depending on their chromosome numbers, parental species always contained 24-25 synaptonemal complexes (Fig. 3a; Fig. 4a) in 852 analysed cells (Supplementary Table S3). Flow cytometry of testicular cell suspensions from parental species revealed the presence of haploid and diploid cells with a major peak corresponding to haploid cells (Supplementary Fig. S5a). Histological analysis showed many sperm cells organized in cysts in all sexual males investigated, containing properly developed Sertoli and Leydig cells (Supplementary Fig. S6a), suggesting normal development of testes.

Sterility of hybrid males is caused by both, the aberrant pairing in the meiotic prophase and the inability to proceed to meiosis

Flow cytometry revealed that the analysed hybrid males lacked the haploid population of cells, characteristic for spermatids and spermatozoa of sexual males. Hybrid males had only diploid cell populations (Supplementary Fig. S5b-h). Interestingly, TB males had a small population of diploid cells corresponding to pachytene stage. These males had morphologically underdeveloped gonads, indicating that here the sterility is probably caused by the inability of gonocytes to proceed to meiosis (Supplementary Fig. S4a,b; Supplementary Fig. S5h). Histological examinations demonstrated that analysed hybrid males, had defective testes with an asynchronous development of germinal cells in cysts (Supplementary Fig. S6b-g). Similar results were reported in naturally occurring ET hybrid males 15.

Furthermore, hybrid males from most experimental crosses showed aberrant chromosomal pairing during the pachytene in 244 analysed cells (Supplementary Table S3). Abnormal cells had variable numbers of univalents, bivalents and multivalents (Fig. 3c; Supplementary Fig. S7a-g). The only exception to aforementioned observations were the NP hybrids involving closely related parental species (C. tanaitica and C. pontica) where we detected fully-paired chromosomes during the pachytene, in 36 cells (Fig. 3b). Chromosomal spreads of NP males also clearly showed the presence of formed spermatid nuclei, similar to sexual males (Fig. 3b). Additionally, we counted SCs, based on the presence of crossing-over foci (Supplementary Fig. S10a-f) 15,31.

Production of unreduced gametes by premeiotic genome endoreplication is the prevalent gametogenic pathway among hybrid females

Analysis of diplotenic oocytes of most hybrid females revealed no univalents or mispaired chromosomes, but instead showed the formation of 49 or 50 properly formed bivalents (Supplementary Fig. S8a-e). Similar to previous data on natural clones 13,29, we therefore conclude that diplotene oocytes in the majority of hybrid females contain only cells with duplicated genomes, suggesting their ability to complete clonal gametogenesis (Fig. 4; Supplementary Fig. S8a-e). The only exceptions were observed in OB and NP hybrids, where diplotenic oocytes contained only 25 fully formed bivalents, suggesting normal meiosis with the pairing likely between orthologous chromosomes (Supplementary Fig. S8f).

However, the analysis of 665 pachytenic oocytes revealed different patterns (Supplementary Table S3). OB and NP females possessed 25 bivalents with delayed pairing of two chromosome pairs (Supplementary Fig. S9d), generally matching their diplotene configurations. However, most F1-females contained pachytenic cells (n = 535) with abnormally-paired chromosomes and various numbers of univalents, bivalents, and multivalents (Fig. 4c; Supplementary Fig. S9a-c, e-h). This suggests that most pachytenic oocytes in hybrids did not undergo genome endoreplication. Pachytenic oocytes with duplicated genomes and thus normal chromosomal pairing were only observed in females from three F1-families (EC, EO and ET hybrids). These showed sets of 50 or 49 bivalents with properly loaded lateral and central synaptonemal components and MLH1 loci on each bivalent (Fig. 4d).

Overall, the incidence of genome endoreplication was very low among pachytenic cells of any F1 hybrid female, occurring on average in 0.7% of cells only (Supplementary Table S3) and there were no significant differences between any combinations of parental species (ANOVA Chi square test of generalized linear mode (GLM); DF = 9, dev. = 7.5, p-value = 0.59). We also did not observe any significant effects of maternal species in those crosses, where both directions were performed, i.e. there were no obvious differences between ET and TE as well as between OE and EO cross types.

We therefore treated all experimental progenies as a single group and tested whether they differed from previously-analysed natural diploid ET-hybrids, whose incidence of duplicated gonocytes was higher, on average ∼6%; 29. Here, the difference was significant (ANOVA Chi square test of GLM; DF = 1; dev = 22.07; p-value < 10-5), suggesting that natural clones produce a higher proportion of oocytes with endoreplicated genomes than experimental F1 hybrids.

The contrasting incidence of oocytes with endoreplicated genomes between pachytene (approx. 6 %) and diplotene stages (100%) is consistent with previous results 29 suggesting that premeiotic genome endoreplication is a rare event in all types of crosses. This allows cells to form normal bivalents and to enter the diplotene, while the majority of cells remain non-duplicated and cannot proceed beyond the pachytene checkpoint due to mispairing among orthologous chromosomes.

Importantly, we noted that incidence of duplicated gonocytes was significantly higher among F1 hybrids (treated as a single group), than in females of any tested parental species, where in fact no duplications have been observed among 792 analyzed oocytes (ANOVA Chi square test of GLM; DF = 1; dev = 7.506; p-value = 0.00615).

Divergence of parental karyotypes affects the formation of bivalents between orthologous chromosomes in hybrids but has different effects in males and females

We than focused exclusively on those pachytenic cells with unduplicated genomes and found that F1-hybrids largely vary in numbers of paired orthologous chromosomes, ranging from 3 detectable synaptonemal complexes in EO males to 25 fully-formed bivalents in OB females. To explain such a variability, we fitted the distribution of numbers of synaptonemal complexes by variables assuming a) sex of the hybrid individual, b) chromosomal divergence between its parental species measured through the AKD index and c) their genetic divergence measured in exome-wide SNP p-distance. All three variables significantly contributed to the data distribution with numbers of bivalents in pachytenic cells being negatively correlated with morphological dissimilarity of parental karyotypes (GLMM z value = -5.86; p < 10-8) and with their genetic distance (GLMM z value = -3.73; p < 10-3). Males also tended to have fewer bivalents than females (z = -4.54; p < 10-5). The model also suggested that both slopes are steeper in males than in females, however, the interaction of AKD or SNP divergence with hybrids’ sex was not significant (Fig. 5). This suggests that the formation of bivalents in hybrids is negatively affected by chromosomal dissimilarity as well as overall genetic divergence between parental species, and that male offspring generally form less bivalents than female progeny.

Effect of karyotype dissimilarity on numbers of bivalents in hybrids.

Boxplots showing the number of synaptonemal complexes per cell (y-axis) of the F1 hybrids ranked along x-axis according to the morphological divergence of karyotypes between parental species expressed by AKD index 28. Numbers indicate the AKD index values, subscripts F and M indicate females and males, respectively and experimental crosses are indicated by respective letter combinations.

Comparison with other crossing experiments within Cobitidae

Finally, to put our results in context with previously published data, we plotted the reproductive outputs of all known loach hybrids in relation to genetic divergence and chromosomal differences between their parental species. To express the amount of chromosomal dissimilarity among parental species, we used the AKD index calculated from published and here newly prepared karyotypes of all spined loaches known to hybridize. To express the amount genetic divergence among these species, we used their sequence distances in the nuclear RAG1 gene, as a proxy, because exome-based sequence data are not available for some of the previously examined crosses. Data on RAG1- and karyotype divergences were then correlated with reproductive outputs in their hybrids and presented on Fig. 6a.

Reproductive outcomes of hybrids from crosses between species from the subfamily Cobitinae (a).

Plot demonstrating the relationship between reproductive outcome and viability of F1 hybrid loaches and the genetic and karyotype divergences among parental species. Data on F1 hybrids have been obtained in this study as well as from the literature. Karyotype divergence is marked as AKD index and genetic distance is estimated from published RAG1 sequences (K2P distance). Green color shows the ranges of hybrids with sexual reproduction; yellow color shows the ranges of F1 hybrids, which produce unreduced gametes; purple color indicates the ranges of sterility in both sexes; red color shows the ranges of inviable hybrids. To rank reproductive outcomes for every type of cross we labeled them with additional symbols: – both sexes producing haploid gametes with fully paired chromosomes; –both sexes are fertile producing haploid gametes, some females produce unreduced gametes; – females produce haploid eggs, males are sterile; – females produce clonal eggs, males are sterile; – both sexes are sterile; – both sexes are inviable. Asterisk (*) indicates cross combinations for which diplotene analysis is missing. Each abbreviation represent a subgenome of the parental species: Cobitis elongatoides = EE; C. taenia = TT; C. tanaitica = NN; C. taurica = CC; C. pontica = PP; C. strumicae = SS; C. bilineata = BB; C. ohridana = OO; C. hankugensis = HH; C. biwae = BW; C. lutheri = Lut; C. tetralineata = Tet; Iksookimia longicorpa = LL; M. anguillicaudatus clade A = MaA; M. anguillicaudatus clade B = MaB; Lefua echigonia = Lef. Hypothetical scheme of the selective processes necessary for the emergence of clonality in F1 hybrids (b). First step includes the accumulation of mutations in gametogenic regulatory genes which is required to induce premeiotic genome endoreplication in F1 hybrid females. Second step includes the chromosomal divergence which leads to the aberrant chromosomal pairing followed by the cell arrest of meiocytes with non-duplicated genomes. Premeiotic genome endoreplication is sex specific, while hybrid males are sterile.

Inviable or fully sterile hybrids were generally found among hybrids between the genetically most distant species, while sexual and/or clonal reproduction were observed in hybrids between more closely related species. However, the window of genetic divergences allowing for sexual as well as asexual reproduction was relatively wide (Fig. 6a). Interestingly, we noticed that asexuality tends to be observed among female hybrids between species with substantially diverged karyotypes, while sexual reproduction generally required more similar karyotypes.

Discussion

Using freshwater fishes, loaches (Cobitidae), as a model, the present study reveals a wide range of reproductive outcomes in F1 hybrids in relation to genetic and karyotype divergences between the parental species (Fig. 6). On one extreme, hybrid inviability has been reported in crosses between distant species 32. On the other extreme, the production of haploid gametes was found in crosses of closely related species, such as between C. lutheri and C. tetralineata 33, and between C. pontica and C. taenia 17 or C. tanaitica species, whose divergence has been dated to ∼1 Mya 17, but also in OB female hybrids in a relatively distant species pair (Fig. 2, Fig. 6, Supplementary Fig. S9d, S10d,). However, the most common types of reproductive outputs were hybrid sterility and production of unreduced gametes.

In case of hybrids between relatively far-distant species pairs, like the intergeneric Cobitis-Misgurnus cross, or between Adriatic and Cobitis sensu stricto lineages (i.e. EB and TB hybrids), sterility occurred in hybrids of both sexes (Fig. 6). However, the most common outcome was asymmetric, with sterility affecting hybrid males, while F1 females produced unreduced oocytes. This pattern occurred in the majority of crosses, including species pairs, like SE, SP or EO, with divergences reaching ∼17 or even >20 Mya back 22 (Fig. 2, Fig. 6) and it has also been reported from several natural hybrid complexes 13,15,29. An intermediate situation was observed in TP-hybrids between a closely related species that diverged ∼1 Mya 17, where males produced reduced gametes and females produced both, reduced and unreduced ones.

It is generally assumed that the emergence of asexuality is a rare phenomenon that may require complex changes in gametogenic pathways and reproductive modes 19. This opinion has been corroborated by the “twiggy” distribution of ‘asexual’ organisms on the tree of life 34 and by many failures to re-generate naturally occurring asexual hybrids by experimental crosses of their parental species (e.g. in apomictic Poecilia fish and parthenogenetic Darevskia lizards 19,35.

However, we observed the production of unreduced gametes in 7 out of 11 crossed species pairs of loaches and in 10 out of 18 species pairs when including literature reports and natural hybrids (Fig. 2, Fig. 6). Such a widespread emergence of asexual gametogenesis therefore appears not to depend on the presence of a genome from any particular parental species, but to arise relatively commonly among independent species pairs across the entire subfamily Cobitinae. Asexual gametogenesis has also been observed in experimental F1 progeny of several other vertebrate species 36. This suggests that the formation of unreduced gametes, i.e. the necessary prerequisite for asexual reproduction, may be a much more common outcome of interspecific hybridization than it would appear from the rarity of asexual hybrid taxa that are successfully established in nature.

Some studies also suggested that the emergence of asexuality requires the evolution of specific genes within hybrid lineages after their origin 37. However, our finding that asexual gametogenesis originates relatively commonly already in the F1 generation somewhat refines this hypothesis and suggests that it is a direct result of genome merging by interspecific hybridization. Importantly, all F1-females capable of clonal gametogenesis employed an identical cytogenetic mechanism, i.e. premeiotic genome endoreplication, as also observed in many natural asexual vertebrates 13,15,16,29,36,38,39 suggesting that gametogenic perturbations induced by hybridization are often canalised into a similar developmental pathway. The reasons for such common patterns are not clear. The regulation of cell cycle is particularly sensitive to stress and perturbations 40 and interspecific hybridization may act as such stimulus 18. While asexual reproduction may employ various pathways 10, it is possible that aberrations of cell cycle leading to endoreplication are more common since they require less accumulated incompatibilities than other routes to the production of unreduced gametes. Alternatively, the occurrence of endoreplication may reflect some predispositions, like the spontaneous production of unreduced gametes in some sexual parental lineages, which would then ensure clonal reproduction in their hybrids. However, our data argue against this alternative, since we analysed hundreds of gonocytes of sexual females without a single incidence of genome endoreplication 15,29. Moreover, during years of crossing experiments and population sampling, we genotyped hundreds of C. elongatoides and C. taenia and their progenies and never found an autotriploid specimen 23,41, which would be expected if fertilisation of unreduced gametes existed in these species 42.

Our in-depth analysis of hybrid gametogenesis in Cobitis loaches further provided compelling evidence for a tight link between hybrid incompatibility, asexuality and sterility. Namely, the majority of meiocytes produced by both the male and female hybrids failed to form a full set of bivalents between orthologous chromosomes and were arrested at meiotic checkpoints (Fig. 3). However, unlike their sterile brothers, hybrid F1 females also usually contained a minor proportion of oogonia that underwent premeiotic genome endoreplication. In most our F1 hybrid females, only these rare oogonia were able to complete meiosis and form diploid gametes. Interestingly, similar observations were reported from several natural and experimental hybrid vertebrates, like ET and EN hybrid loaches, geckos and whiptail lizards (Aspidoscelis) Medaka ricefish (Oryzias) 16,29,36,43. This suggests that even successfully established hybrid asexuals usually contain only a minor proportion of oogonia, which actually undergo clonal gametogenetic pathway.

Such congruent patterns across phylogenetically distant taxa thus conform to the concept of an “extended speciation continuum”. Similar to the classical concept of the “speciation continuum”, it assumes that speciation generally proceeds from early stages characterized by the production of sexual hybrid progeny to irreversibly diverged species, whose hybrids are either sterile or even inviable. At intermediate stages, however, hybrids between diverging evolutionary lineages may not only have a certain incidence of sterile gonocytes, but also tend to produce unreduced gametes (Fig. 6). Genome endoreplication thus appears as an effective way to overcome sterility in hybrid females, even at high levels of divergence. Asexuality may thus often represent the only reproductive pathway of hybrids between a given pair of species and, together with hybrid sterility, it may simultaneously contribute to reproductive isolation of parental species due to clonal genome propagation.

We observed that the fertility of female hybrids remains conserved due to sex-specific emergence of clonality, while males of the same cross are usually sterile. This may potentially indicate some analogy to other classical sex-related asymmetries in hybrids, like incompatibilities between sex chromosomes (Haldane’s rule) or other uniparentally inherited factors (Darwin’s corollary 45). However, the sex determination systems are not yet known in most of the investigated species 9 and in any case, the reasons for such an asymmetry are likely more complex. For instance, meiocytes in hybrid males contained significantly lower numbers of properly-formed synaptonemal complexes than non-duplicated meiocytes in females (Fig. 5) and were more severely affected by decreasing similarity of parental karyotypes (see Results, 2.3.5.). This indicates some more fundamental differences in pairing mechanisms and affinity among orthologous chromosomes fundamentally between spermatogenesis and oogenesis, e.g. 46. In fact, we recently found that spermatogonial cells taken from a sterile ET hybrid male may resume the ability of endoreplication when transplanted into female gonads 41. This suggests that initiation of premeiotic endoreplication may rely on female-specific gonadal environment rather than on genetic sex determination.

The genetic divergence among parents may affect early gametogenic stages in hybrids through various mechanisms causing e.g. depletion of primordial germ cells 47, improper pairing of diverged chromosomes 48, asynapsis between sex chromosomes 49, or deleterious epistatic interactions between orthologous genes and their binding motifs on chromosomes 50. In loaches, we documented at least two such mechanisms underlying hybrid sterility. TB and PS hybrid males had almost no spermatocytes, suggesting their gonocytes probably failed the transition to meiosis. The other type of sterility conforms to a chromosomal speciation model 51,52, prevailing in most F1 hybrid males, in which spermatocytes developed but possessed aberrantly-paired chromosomes, preventing further progression to spermatids. Analogous patterns were observed in non-duplicated oocytes of most hybrid females, suggesting that problems in bivalent-formation have a detrimental impact on the meiosis in both male and female loach hybrids.

Karyotype evolution among parental species indeed plays a crucial role in hybrids’ reproductive capabilities 52. Our data suggest that it has a crucial, yet indirect, impact on asexual reproduction. Specifically, hybrids between species with diverged karyotypes had less bivalents in non-duplicated meiocytes than those combining similar karyotypes. Sometimes, even hybrids between phylogenetically distant, yet karyotypically similar species could produce full sets of bivalents (e.g. OB hybrids) (Fig. 5). On the other hand, the numbers of endoreplicated oocytes were low in any hybrid. Their proportion was relatively higher in natural clones than in experimental F1 Cobitis, suggesting that such a trait may recover during the evolutionary establishment of a natural asexual lineage 29 and/or that selection among “freshly” emerged clones would favour those with the highest proportion of endoreplicated oocytes. Nevertheless, even the most widespread clones possessed only ∼6% of such oogonia, suggesting that karyotype similarity allows the completion of meiosis and hybrids would mostly reproduce sexually by reduced gametes, even if they had the capacity of producing unreduced gametes, too.

While hybridization is assumed to present an important trigger of asexuality in vertebrates and other organisms 10, the origins of clonal lineages are thought to be rare, and often can be traced back to a single or few hybridization events in a given species pair 19,28. Multiple independent origins of clonal gametogenesis in spined loaches 23,28, this study) in part contradicts this view since interspecific hybridization has been found to result in clonal gametogenesis relatively commonly up to a substantial divergence, where fertile or viable hybrids are no longer appearing (Fig. 6a). However, we emphasize that successful establishment of hybrid clones under natural conditions is likely restricted by several additional levels of selective pressure and thus filtering (Fig. 6b). On the one hand, the “balance hypothesis” 18 and its expansions and fine-tuned versions 17,19, suggest that the production of unreduced gametes probably requires some level of genetic divergence among the parental species to alter the hybrid cell cycle towards genome endoreplication. On the other hand, it appears that most meiocytes of asexual hybrids anyway do not pass through endoreplication 16,29,36,43. Consequently, if the parental karyotypes remain largely similar, then orthologous chromosomes may successfully pair even in such nonduplicated meiocytes, potentially, allowing them to accomplish normal meiosis and form reduced gametes. This would decrease the proportion of unreduced gametes produced by such hybrids and negatively impact on their likelihood to establish a stable asexual lineage.

We therefore propose, that for a cross to be successful in generating an asexual hybrid, it not only has to produce a sufficient number of unreduced gametes, but it also has to prevent the formation of reduced gametes from non-duplicated meiocytes. One way to achieve this may be if hybridization involves species with sufficiently diverged karyotypes, ensuring improper pairing of orthologous chromosomes in non-duplicated gametes. Increased divergence of parental karyotypes may therefore have a positive impact on the production of stable clonal hybrids and, for instance, the fixation of certain chromosomal rearrangements among diverging populations may be accelerated by certain geographical aspects of speciation, such as parapatry 51,52. It would thus be attractive to test if the incidence of asexual hybrids correlates with increased rates of karyotype evolution among parental species.

The rarity of naturally occurring clonal organisms probably results from the simultaneous impact of the abovementioned selective processes. However, our data indicate that the production of unreduced gametes, which is the very basic conditio sine qua non for the evolution of asexuality, may be surprisingly “easy” to trigger, especially in hybrids of distantly related lineages. Of course, we may not rule out that the frequent emergence of premeiotic endoreplication in Cobitidae reflects some specific predispositions in their gametogenic machineries in comparison to other taxa, where asexuality appears rarer. Thus, our study demonstrates that conclusions about the evolution of asexuality should not only be drawn from natural asexuals, but should be based also on experimental crosses.

Supporting information